---
title: Advent of Code 2021 - Day 1
author: admin
date: '2021-12-01'
slug: []
categories:
  - R
tags:
  - R
  - Advent of Code
subtitle: ''
summary: ''
authors: []
lastmod: '2021-12-02T18:42:43+01:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

Hi,

welcome to my solutions to the challenges of *Advent of Code 2021*. I present my solutions as a short post for every daily challenge. If you do not know what is going on, please go ahead and check out <https://adventofcode.com/>. I will not go into detail about the problem for day one, since you can simply visit <https://adventofcode.com/2021/day/1>.

The goal for this year is to solve every puzzle. The puzzles should get harder every day until the 24th of december. I am going to use *R*, with no added packages. Please the my setup below.

```{r session-info}
sessionInfo()
```



## Solution to day 1

```{r day-one-one}
# Data --------------------------------------------------------------------
sonar_data <- read.csv(file = "day_one_sonar",  col.names = "depth", header = FALSE)
head(sonar_data)
```

Part 1:

This can be solved using **diff** and summing over the logical vector, where **diff_depth > 0**.

```{r}
# Solution -----------------------------------------------------------------
diff_depth <- diff(sonar_data$depth)
sum_inc <- sum(diff_depth > 0)
sum_inc
```

Part 2:

I solved part two using a **for** loop, though there might be a simpler solution.

```{r day-one-two}
# Solution -----------------------------------------------------------------
shift_sum <- vector(mode = "numeric", length = nrow(sonar_data) - 2)
for(i in seq_along(shift_sum)) {
  shift_sum[i] <- sum(sonar_data[i : (i + 2), "depth"])
}

sum(diff(shift_sum) > 0)
```


Yes, it is that easy! See you at day 2!
