---
title: Reading Data - data.table & arrow
author: admin
date: '2021-02-17'
slug: []
categories: []
tags:
  - data.table
  - arrow
subtitle: ''
summary: ''
authors: []
lastmod: '2021-02-19T15:12:05+01:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

Hi,
I recently stumbled upon the discussion on twitter, concerning the speed of reading data into R. A recent conference talk at rstudio::global, where it was claimed that `arrow` could outperform `data.table´s fread` function and the following debate on twitter got me interested in testing `arrow´s` data reading capabilities vs `data.table´s`. This is not a comprehensive benchmark, rather a test for data I used in one of my last project. 

## Data

We are going to download water level data from Hamburg, St.Pauli, Germany.
You can access the data here: <https://www.kuestendaten.de/DE/dynamisch/Funktionen/Liste_der_vorhandenen_Daten/index.php.html>

You need to input a few parameters here:

- Portal                 = Portal Tideelbe
- Erhebende Organisation = Hamburg Port Authority (HPA)
- Themenbereich          = Alle
- Einzelthema            = Wasserstand

Then choose the data for St. Pauli (*Wasserstand (Ganglinie)*)
The downloaded file is a zip file, which you can unzip using something like (using your path to the downloaded data):

```{r eval=FALSE, include=TRUE}
unzip(zipfile = file.path("path"), exdir = file.path("path"))
```

You should get a txt file named *directDownload_97.txt*, which we are going to read into R using the functions `data.table::fread` and `arrow::read_delim_arrow`. To make things a little bit easier I deleted the header and the tail of the file by hand.

## A first try

Now we load the packages into our R session and set the path to the file you created with `unzip` and cleaned by hand.

```{r message=FALSE, warning=FALSE}
library(data.table)
library(arrow)
library(microbenchmark)
```

You need set the path to the unzipped and cleaned file.

```{r}
path_to <- file.path("D:", "Projekte", "TestsPackages", "data", "directDownload_97.txt")
```

We define two functions, which are calling the `arrow` or `data.table` functions.

```{r}
#Calling arrow
arrow_read <- function() {
  read_delim_arrow(file = path_to, delim = "\t", 
                   col_names = FALSE)
}
#Calling data.table
dt_read    <- function() {
  fread(file = path_to, sep = "\t", header = FALSE)
}
```

Using the `microbenchmark` package, we call the functions five times each and print/plot the results. Each function should be called 3 times as a block. 

```{r}
mb <- microbenchmark::microbenchmark(dt_read(), arrow_read(),  
                                     times = 3, control = list(order = "block"))
```

```{r}
print(mb)
```

```{r default_classes_plot, echo=FALSE}
plot(mb, main = "arrow vs. data.table - default classes")
```

So, `arrow::read_delim_arrow` is super fast here, comparable do `data.table::fread`. We take a look at the objects that are being created.

```{r}
da <- arrow_read()
str(da)
```


```{r}
dt <- dt_read()
str(dt)
```

`arrow_read()` creates a `tibble` with three columns (f0, f1, f2). With f0 being a `POSIXct` vector, while `dt_read()` returns a `data.table`, with columns V1, V2, V3. 

## Setting column classes

We overwrite the functions setting the column classes explicitly.

```{r}
#Calling arrow
#setting the col_types parameter
arrow_read <- function(col_classes = c("tii") ) {
  read_delim_arrow(file = path_to, delim = "\t", col_types = col_classes, 
                   col_names = c("V1", "V2", "V3"))
}
#Calling data.table
#setting the colClasses parameter
dt_read    <- function(col_classes = c("POSIXct", "integer", "integer")) {
  fread(file = path_to, sep = "\t", header = FALSE, tz = "UTC",
                    colClasses = col_classes)
}

```

`arrow` and `data.table` automatically detect that the first column is a date/time object and reads it as `POSIXct`, whereas Let us assume for now that the first column should be read as a character vector. We can set the parameter of the functions as such:


```{r}

a_c <- c("cii")
d_c <- c("character", "integer", "integer")

mb <- microbenchmark::microbenchmark(arrow_read(col_classes = a_c), 
                                     dt_read(col_classes = d_c), 
                                     times = 3, control = list(order = "block"))
```

The results of the benchmark is shown below:

```{r echo=FALSE}
print(mb)
```

```{r character_column_plot, echo=FALSE}
plot(mb, main = "arrow vs. data.table - character column")
```

## Threads

Since I am new to the `arrow` library, I do not know, how many threads it uses. We can check and set the number of threads for `arrow` and `data.table` using built-in functions.

```{r}
arrow::cpu_count()
```

```{r}
data.table::getDTthreads()
```

This makes we wonder what happens when we set the threads and compare the reader functions again.

```{r}
SetThreads <- function(n_threads = 2L) {
  arrow::set_cpu_count(num_threads = n_threads)
  data.table::setDTthreads(threads = n_threads)
}
```


```{r}
mb <- vector(mode = "list", length = 4L)

for(i in 1L:4L) {
  SetThreads(n_threads = i)

a_c <- c("cii")
d_c <- c("character", "integer", "integer")

mb[[i]] <- summary(microbenchmark::microbenchmark(arrow_read(col_classes = a_c), 
                                     dt_read(col_classes = d_c), 
                                     times = 3, control = list(order = "block")))

}

mb <- rbindlist(mb, idcol = "threads")
```
```{r}
print(mb)
```


## Wrapping up

I use data.table on a daily basis for my scripts and frequently call `fread` to read data into my R session. `arrow` is super impressive and performance of both functions is amazing. Thanks to all package developers for enhancing our lifes :-). See you soon!
