---
title: data.table - useR!2019 I
author: admin
date: '2019-07-14'
slug: data-table-new-and-old-functions
categories: []
tags:
  - data.table
  - useR!2019
subtitle: ''
summary: ''
authors: [admin]
lastmod: '2019-08-27T13:40:04+02:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

This is the first post looking at the **data.table** presentations from **useR!2019** which was held in Toulouse from 2019-07-07 to 2019-10-10. For starters we are going to take a look at the presentations from *Arun Srinivasan*, one of the main developers of **data.table**. His talk is called **THE #RDATATABLEPACKAGE
for fast, flexible and memory efficient data wrangling**. 

Talk Agenda:

- General data.table syntax
- Explaining .SD & .SDcols
- optimisation & new functionalities

## General data.table syntax

Following code chunk installs the development version of the package, which is needed for some functions explained below.

```{r eval=FALSE, include=TRUE}
install.packages("data.table", repos="https://Rdatatable.gitlab.io/data.table")
```


We start of with a simple expample.

```{r}
library(data.table)
```

We define a data.table DT using the `fread` function.

```{r}
(DT <- fread("id code valA valB
1 c 1 10
1 b 2 11
1 c 3 12
1 c 4 13
2 a 5 14
2 a 6 15
1 b 7 16
2 a 8 17
1 c 9 18"))
```

The general data.table syntax is `DT[i, j, by]`, which you can translate into: *i = on which rows, j = what to do (columns), grouped by what*. So when we want to get the sum of *valA*, we can simply do.

```{r}
DT[, sum(valA)]
```
We have not defined a *i* nor *by* argument here. This simply means we want to compute for all rows and grouping is not relevant. Now we group by id.

```{r}
DT[, sum(valA), by = id]
```
We can also name the outcome of the column directly. The `.()` is an alias for `list`.

```{r}
DT[, .(sumA = sum(valA)), by = id]
```
When you pass a column to the `by` argument, internally the number of the unique values for that column gets evaluated. Afterwards the rows are associated to the respective group. Then the expression provided in `j`, `sum(valA)` gets evaluated for the groups. 

This is a very simple example, but illustrates the general form quite nicely. *Arun* introduces another simple example by passing an expression to the *i* argument. 

```{r}
DT[code != "b", .(sumA = sum(valA)), by = id]
```

Here we simply state that `code` should be not equal to "b". Very easy and logical syntax.


## Explaining .SD & .SDcols

We are still operating on DT.

```{r}
DT
```

`.SD` stands for *Subset of Data*. Together with `.SDcols` we can use it to simply subset `DT` by column names and then compute on this subset.


You could do this:


```{r}
DT[, .SD, .SDcols = "valA"]
```

Which returns a `data.table` object with one column `valA`. You can of course pass more columns.

```{r}
DT[, .SD, .SDcols = c("valA", "valB")]
```

you can also use the `patterns` function in `.SDcols`.

```{r}
DT[, .SD, .SDcols = patterns("^val")]
```

How can we use this functionality to compute the sum of `valA` and `valB` grouped by `id` where `code != "b` ?
Easy :-)

```{r}
DT[code != "b", lapply(.SD, sum), .SDcols = patterns("^val"), by = "id"]
```
The `patterns` function only returns columns starting with *val*. 
`.SD` contains all the columns, except for the grouping columns supplied in `by`, for each group.

```{r}
DT[code != "b", print(.SD), .SDcols = c("code","valA", "valB"), by = id]
```
Once you grouped by `id` only the columns `valA` and `valB` are being considered due to `.SDcols`. We therefore compute the sums on on these columns and not on `code`.

```{r}
DT[code != "b", print(.SD), .SDcols = c("valA", "valB"), by = id]
```

So how does this `lapply(.SD, sum)` work? `lapply` is the infamous base function which returns a list. We operate on `.SD` and compute the sum of every column in `.SD`. We take a step back and look at the `class` and `typeof` of `DT`

```{r}
class(DT)
```


```{r}
typeof(DT)
```


```{r}
typeof(DT[,.SD])
```
You find more information about `.SD` in this *stackoverflow* post:

<https://stackoverflow.com/questions/8508482/what-does-sd-stand-for-in-data-table-in-r>

You can also check `vignette("datatable-sd-usage")` in the development version.

## Optimisation

Arun presents optimisation principles for operations in `i`, `j` & `by`. 

### Optimisation in "i" 

We create a `data.table` which has 200 million rows and two columns `x` and `y`.

```{r eval=FALSE, include=TRUE}
dt <- data.table(x = sample(1e5, 2e8, TRUE), y = runif(2e8)) 
```

Now we only return rows where the value of `x` is in `1000:2000`, we operate in `i`. 

```{r eval=FALSE, include=TRUE}
system.time(dt[x %in% 1000:2000])
system.time(dt[x %in% 1000:2000])
```

This takes a while (~17s on my laptop), but running the expression a second time results in a substantial speed up (~4s total). Please note that this not a proper benchmark. The second time we call this the expression, it reuses the index created in the first run. This currently works for `==` and `%in%` and is called *Auto indexing*. The index is stored within the `data.table` as an attribute.


Now we create a `data.table` with 20 columns and 50 million rows to demonstrate that expressions in `i` can run in parallel.

```{r eval=FALSE, include=TRUE}
dt <- setDT(lapply(1:20, function(x) sample(100, 5e7, TRUE))) 
```

We operate in `i` to return the rows, where `V1 > 50L`. The column is processed in parallel.

```{r eval=FALSE, include=TRUE}
system.time(dt[V1 > 50L])
```

You can set the threads `data.table` uses with and check the differences.

```{r eval=FALSE, include=TRUE}
setDTthreads(threads = 1L)
system.time(dt[V1 > 50L])
setDTTthreads(threads = 2L)
```


### Optimisation in "by"

Recently the radix order has been parallelised. We define a new data set (200 million rows, 2 cols, ~3GB).

```{r eval=FALSE, include=TRUE}
dt <- data.table(x=sample(1e5, 2e8, TRUE), y=runif(2e8))
```

Now we run this expression, which returns the occurences grouped by `x`.

```{r eval=FALSE, include=TRUE}
dt[, .N, by = x]
```


### Optimisation in "j"

In `j` certain functions are optimized. Arun mentions `mean`, `median`, `sum`, `min`, `max`, `head`, `tail`. 
We can see a difference, when comparing `base::mean` with just calling `mean`in `j`.

```{r eval=FALSE, include=TRUE}
dt[, lapply(.SD, base::mean), by=V1]
```


```{r eval=FALSE, include=TRUE}
dt[, lapply(.SD, mean), by=V1]
```

Internally this expression uses the function `mean` implemented by `data.table` in *C*. 

## New functionalities

The first function we are going to look at is `nafill`. This function allows us to fill `NA` values in a `data.table` by a specific method, like *last observation carried forward (locf)*. See `?data.table::nafill` for more information.

### nafill

`nafill` is a function that fills missing values in a `data.table`. We define `DT` as follows. 

```{r}
(DT <- fread("V1 V2 V3 V4
1 c NA 10
2 b 2 NA
NA c 3 NA
1 NA 4 NA
2 NA 5 14"))
```

Now let us try to apply `nafill`to this object. The function takes in three important arguments:

```{r eval=FALSE, include=TRUE}
nafill(x, type=c("const","locf","nocb"), fill=NA,
       verbose=getOption("datatable.verbose"))
```

`x` can be a vector, list, data.frame or data.table. The `type` defines how the missing values should be filled.

- `locf` -> last observation carried forward
- `nocb` -> next observation carried backwards
- `const` -> a constant

```{r}
#(DT <- nafill(DT, "locf"))
```

Above code will throw an error indicating that we can only fill columns with type `numeric` (double & integer). Here the `setnafill` functions comes in handy:

```{r}
#getting numeric cols

numeric_cols <- names(DT)[DT[, sapply(.SD, is.numeric)]]

setnafill(DT, "locf", cols = numeric_cols)
DT[]
```

This updates `DT` by reference.  


### frollmean & frollsum

Now we are going to explore the `frollmean` and `frollsum` functions. These functions allow us to compute rolling means and rolling sums. 

```{r}
DT[, lapply(.SD, frollmean, 3), .SDcols = numeric_cols]
#see also ?frollsum
```

```{r}
DT[, lapply(.SD, frollsum, 3), .SDcols = numeric_cols]
```


### coalesce

The `coalesce` function fills missing values in a vector by trying to pull values from one or more candidate vectors. 

```{r}
x <- c(11L, NA, 13L, NA, 15L, NA)
y <- c(NA, 12L, 5L, NA, NA, NA)
z <- c(11L, NA, 1L, 14L, NA, NA)

coalesce(x, y, z)
```

Given above syntax `coalesce` tries to fill missing values in `x` by looking at the corresponding elements of `y` & `z`. At the second index we find a missing value in `x`, whereas `y = 12L ; z = NA`. At the fourth index we find `y = NA ; z = 14L`. So the new vector becomes `11 12 13 14 15 NA`.

## Helpful links

-The talk: <https://www.youtube.com/watch?v=tWx1ooBSxFc>
- Slides: <http://www.user2019.fr/static/pres/t258038.pdf>
- github: <https://github.com/Rdatatable>, <https://github.com/Rdatatable/data.table/wiki>


See you soon!
