---
title: data.table - Generate and fill missing values.
author: ''
date: '2019-11-04'
slug: data-table-cross-joining-i
categories: []
tags:
  - data.table
  - cross join
  - time series
  - missing values
subtitle: ''
summary: ''
authors: [admin]
lastmod: '2019-11-04T15:55:01+01:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

Hi,

in this post I illustrate how we can generate `NA` (missing values) rows using only `data.table` functions, where you have a *date* column (or equivalent) in your data.  
We begin by loading the package.

```{r}
library(data.table)
```

# Problem 

Let us assume you have a `data.table` with three columns `date`, `value` and `product`. Here `date` is of type *IDate*, `value` & `product` are *integer*. You can think of it as sales data, where you observe daily sales for a certain product. Thus giving you a time series for every product. We assume that we only get a new entry, when we actually observe a sale. Let us look at a minimal example.

```{r}
(dt <- data.table(date = as.IDate(c("2019-11-01","2019-11-02", "2019-11-04", "2019-11-07")), value = c(3L,4L,1L,8L), product = c(1L,2L,1L,1L)))
```

```{r}
sapply(dt, typeof)
```

For the `date` column you get `integer`, because the `IDate` class is a integer based date class. Please check `?IDate` for more information.

Since we want a row for every possible date within a certain range, we need to insert rows, where we did not observe anything.

# Solution

It is obvious that we have not sold anything on certain days. Now, our job is to insert rows where we write `NA` for `value` by `date` and `product`. There are different approaches for this task, we focus on joining. Since we know the start and the end of the time series, we can generate a sequence.

```{r}
(seq_dates <- seq(as.IDate("2019-11-01"), as.IDate("2019-11-07"), by = "days"))
```

Now we can use the `CJ` function from `data.table` to generate a `data.table`. We join `dt` to the return value of `CJ` based on the columns `date` and `product`

```{r}
dt[CJ(product = unique(product), date = seq_dates), on = c("date", "product")]
```

The `CJ` function simply generates a `data.table` with all possible combinations of `date` and `product`.

```{r}
CJ(date = seq_dates,  product = unique(dt$product))
```


# Scalabilitie

We check the perfomance of the proposed workflow by generating a `data.table` with two parameters: `n_days` & `n_products`. The code below generates random `n_days` within the year 2019 by `product`. The `value` is simply a random integer. This results in an object with 2.5 million rows and three columns.


```{r}
seq_dates  <- seq(as.IDate("2019-01-01"), as.IDate("2019-12-31"), by = "days")
n_days     <- 250L
n_products <- 10000L

dt_large <-data.table(rbindlist(lapply(1:n_products, function(x) {
  date    <- sample(seq_dates, n_days)
  product <- rep.int(x, n_days)
  return(list(date = date, product = product))
})), value = sample.int(10, n_products*n_days, replace = TRUE))
```

```{r}
sapply(dt_large, typeof)
```


Using `CJ` we get `dt_large_na` with 3.65 million rows (`365 days * n_products`) and missing values for `value`, when did not sold a product on a certain day.

```{r}
system.time(dt_large_na <- dt_large[CJ(product = unique(product), date = seq_dates), on = c("date", "product")])
```

```{r}
dt_large_na
```

For product 1 & 5 the data looks like this:

```{r}
dt_large_na[product == 1]
```


```{r}
dt_large_na[product == 5]
```

# Filling missing values

We generated a `data.table` with missing values by joining. There are a lot of techniques to fill `NA` values. I want to quickly demonstrate how you can use the `nafill` function included in `data.table`. I introduce a new column in `dt_large_na` to demonstrate the functionality.

```{r}
dt_large_na[, filled_value := nafill(value, "locf"), by = "product"]
```

This simply fills the `NA` values on a product level by *last observation carried forward (locf)*. You might want to use a more sophisticated approach for real data. Please note that `nafill` only accepts numeric vectors.

```{r}
dt_large_na
```

Imputing missing values requires a lot more thought for real applications. A great resource is: <https://rmisstastic.netlify.com/how-to/impute/missimp>. 

See you soon,

Moritz
